---
title: "CyclisticTripData"
author: "Tom C"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```



## R Markdown for Cyclistic Case Study

This is my first crack at creating a Rstudio and Rmarkdown case study.  

I will start the Rmarkdown documentation at the point where my data is loaded and ready for
data cleaning.

##Install and load packages

```{r}
library(dplyr)
library(tidyverse)  #helps wrangle data
library(lubridate)  #helps wrangle date attributes
library(ggplot2)  #helps visualize data
library(readr)

```

## View the combined_trip_data

###View the dataset headers and find the variable type
###Inspect the new table that has been created
```{r}



combined_trip_data<-read_csv("combined_trip_data.csv")

head(combined_trip_data)
glimpse(combined_trip_data)
colnames(combined_trip_data)  #List of column names
nrow(combined_trip_data)  #How many rows are in data frame?
dim(combined_trip_data)  #Dimensions of the data frame?
head(combined_trip_data)  #See the first 6 rows of data frame.  Also tail(all_trips)
str(combined_trip_data)  #See list of columns and data types (numeric, character, etc)
summary(combined_trip_data)  #Statistical summary of data. Mainly for numerics
```

##Add columns that split out the dates so we can do more analysis
```{r}
combined_trip_data$date <- as.Date(combined_trip_data$started_at)
combined_trip_data$month <- format(as.Date(combined_trip_data$date), "%m")
combined_trip_data$day <- format(as.Date(combined_trip_data$date), "%d")
combined_trip_data$year <- format(as.Date(combined_trip_data$date), "%Y")
combined_trip_data$day_of_week <- format(as.Date(combined_trip_data$date), "%A")
```

##Calculating the ride length and making it numeric so we can do calcs
```{r}
combined_trip_data$ride_length <- difftime(combined_trip_data$ended_at,combined_trip_data$started_at)
str(combined_trip_data$ride_length)
combined_trip_data$ride_length <- as.numeric(as.character(combined_trip_data$ride_length))
is.numeric(combined_trip_data$ride_length)
str(combined_trip_data$ride_length)
```

##remove instances where ride length was negative due to service being performed and create a new copy of the data.
```{r}
new_combined <- combined_trip_data[!(combined_trip_data$start_station_name == "HQ QR" | combined_trip_data$ride_length<0),]
```

##Saving the data
```{r}
save(new_combined,file="new_combined.RData")

```
##removing NA from the dataset and saving as new_combined2
```{r}
new_combined2<-na.omit(new_combined)
save(new_combined2,file="new_combined2.RData")
```
##Looking at key parts of the data-could use summary() but decided manually
##reviewing would be better practice.
```{r}
mean(new_combined2$ride_length)
median(new_combined2$ride_length)
max(new_combined2$ride_length)
min(new_combined2$ride_length)
```

##start comparing members with casual users using aggregate function
```{r}
aggregate(new_combined2$ride_length~new_combined2$member_casual,FUN=mean)
aggregate(new_combined2$ride_length~new_combined2$member_casual,FUN=median)
aggregate(new_combined2$ride_length~new_combined2$member_casual,FUN=max)
aggregate(new_combined2$ride_length~new_combined2$member_casual,FUN=min)
```
##Average ride time for each day of the week
```{r}
aggregate(new_combined2$ride_length~new_combined2$member_casual+new_combined2$day_of_week,FUN=mean)
```

##reorder the days of the week
```{r}
new_combined2$day_of_week<-ordered(new_combined2$day_of_week,levels=c("Sunday","Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
aggregate(new_combined2$ride_length~new_combined2$member_casual+new_combined2$day_of_week,FUN=mean)
```


##average ride time for casual and member each day
```{r}
aggregate(new_combined2$ride_length~new_combined2$member_casual+new_combined2$day_of_week,FUN=mean)
```

##counting number of rides by membe and casual and day
```{r}
new_combined2%>%
  mutate(weekday=wday(started_at,label=TRUE))%>%
  group_by(member_casual,weekday)%>%
  summarise(number_of_rides=n()
  ,average_duration=mean(ride_length))%>%
  arrange(member_casual,weekday)
```

##visualization for number of rides
```{r}
new_combined2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
  ggsave("number_of_rides_plot.png")
```

##visualization for average duration
```{r}
new_combined2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
  ggsave("number_of_rides_plot.png")
```

##export for use in tableau or other software-saving my csv file
```{r}
counts <- aggregate(new_combined2$ride_length ~ new_combined2$member_casual + new_combined2$day_of_week, FUN = mean)
write.csv(counts, file = 'C:/Users/ltcas/Documents/Data Analytics/2021_2022_Trip_Data/trip_data-file/avg_ride_length.csv')
```



























Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
